---
title: "PML project"
author: "mohamed morsy"
date: "June 26, 2018"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data
The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Goal
The goal of your project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## Libraries
```{r}
library(caret)
library(rattle)
library(randomForest)
library(gbm)
```

## Loading data
```{r}
TrainData <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"),header=TRUE)
dim(TrainData)
TestData <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"),header=TRUE)
dim(TestData)
```


## Clean the data

Many columns have NA values or blank values for almost every observation.
The first seven columns give information about the people who did the test, and also timestamps. We will not take them in our model.
```{r}
# train data
indColToRemove <- which(colSums(is.na(TrainData)|TrainData=="")>0.9*dim(TrainData)[1]) 
TrainDataClean <- TrainData[,-indColToRemove]
TrainDataClean <- TrainDataClean[,-c(1:7)]
dim(TrainDataClean)

# test data
indColToRemove <- which(colSums(is.na(TestData)|TestData=="")>0.9*dim(TestData)[1]) 
TestDataClean <- TestData[,-indColToRemove]
TestDataClean <- TestDataClean[,-1]
dim(TestDataClean)
```

```{r}
set.seed(12345)
inTrain1 <- createDataPartition(TrainDataClean$classe, p=0.75, list=FALSE)
Train1 <- TrainDataClean[inTrain1,]
Test1 <- TrainDataClean[-inTrain1,]
dim(Train1)
dim(Test1)
```


we will test 3 different models :  
* classification tree 
* random forest 
* gradient boosting method

```{r}
trControl <- trainControl(method="cv", number=5)
model_CT <- train(classe~., data=Train1, method="rpart", trControl=trControl)
fancyRpartPlot(model_CT$finalModel)
```


```{r}
trainpred <- predict(model_CT,newdata=Test1)

confMatCT <- confusionMatrix(Test1$classe,trainpred)

# display confusion matrix and model accuracy
confMatCT$table
confMatCT$overall[1]
```

The accuracy of this first model is low, about 55%

## Random Forest

```{r, cache=TRUE}
model_RF <- train(classe~., data=Train1, method="rf", trControl=trControl, verbose=FALSE)
print(model_RF)
plot(model_RF,main="Accuracy of Random forest model by number of predictors")
```
```{r}
trainpred <- predict(model_RF,newdata=Test1)

confMatRF <- confusionMatrix(Test1$classe,trainpred)

# display confusion matrix and model accuracy
confMatRF$table
confMatRF$overall[1]
names(model_RF$finalModel)

model_RF$finalModel$classes
plot(model_RF$finalModel,main="Model error of Random forest model by number of trees")
```

```{r}
MostImpVars <- varImp(model_RF)
MostImpVars
```

we reach an accuracy of 99.3% using cross-validation

## radient boosting
```{r}
model_GBM <- train(classe~., data=Train1, method="gbm", trControl=trControl, verbose=FALSE)

print(model_GBM)
plot(model_GBM)

trainpred <- predict(model_GBM,newdata=Test1)

confMatGBM <- confusionMatrix(Test1$classe,trainpred)
confMatGBM$table

confMatGBM$overall[1]
```

## Conclusion
This shows that the random forest model is the best one. We will then use it to predict the values of classe for the test data set.

```{r}
FinalTestPred <- predict(model_RF,newdata=TestDataClean)
FinalTestPred
```

